---
title: "Confidence Intervals for a Proportion"
author: "Devan Becker"
date: "15/07/2020"
output: 
    beamer_presentation:
        colortheme: Western2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(2112)
```

## Introduction

### This is the same as the last lesson

- Based on our data, we make an interval that we think describes the population.\lspace\pause
- In this case, we just have a different population distribution?

### Assumptions

In stats, assumptions give us power, \pause but only if they're good assumptions.

\pause

Assumptions for a CI for $p$ are the same as the assumptions for the binomial distribution, with the addition of an SRS.

## The CI for $p$

### Sampling Distribution of $\hat p$

As we saw before the midterm, if the population is $B(n,p)$, then under certain conditions,

$$\hat p \sim N\left(p, \sqrt{\frac{p(1-p)}{n}}\right)$$


### Deja-Vu

Since $\hat p \sim N(p, \sqrt{\frac{p(1-p)}{n}})$, 

$$
\frac{\hat p - p}{\sqrt{p(1-p)/n}} \sim N(0,1)
$$

\pause Again, we can use the form $z = (x-\mu)/\sigma$, but replace $x$, $\mu$, and $\sigma$ with the correct values.\pause

A $(1-\alpha)$CI for $p$ is:

$$
\hat p \pm z^*\sqrt{\frac{p(1-p)}{n}}
$$

### We don't know the variance, why not $t_{n-1}^*$?

- We used $t_{n-1}^*$ because we had to estimate $\sigma$\newline

- There's no $\sigma$ to estimate!\newline

- The variance of the Binomial distribution is entirely determined by $p$!
    - Binom be crazy.


### ... but Devan, we still don't know $p$!


The $(1-\alpha)$CI for $p$ is:

$$
\hat p \pm z^*\sqrt{\frac{p(1-p)}{n}}
$$
which needs $p$ in the second part of the equation.\pause

\quad

Why not just plug in $\hat p$?\pause

Okay fine. \pause

$\sqrt{\hat p(1-\hat p)/n}$ is called the **estimated standard error**, since its the sd of the sampling distribution, but it's based on an estimate.

### Final_Version_V2_Update_LastTry_Srsly.docx.pdf

The $(1-\alpha)$CI for $p$ is:

$$
\hat p \pm z^*\sqrt{\frac{\hat p(1-\hat p)}{n}}
$$

where $z^*$ is chosen such that $P(Z < -z^*) = \alpha/2$.


### Devan Style: Simulation

```{r}
n <- 100
p <- 0.7
SE_true <- sqrt(p*(1-p)/n)
p_does <- c()
phat_does <- c()
that_does <- c()
z_star <- abs(qnorm(0.05/2))
t_star <- abs(qt(0.05/2, df = n-1))
```

### Devan Style: Simulation

```{r}
for(i in 1:10000){
    new_sample <- rbinom(n=1, size=n, prob=p)
    phat <- new_sample/n
    SE_est <- sqrt(phat*(1-phat)/n)
    
    pCI <- phat + c(-1,1)*z_star*SE_true
    phatCI <- phat + c(-1,1)*z_star*SE_est
    thatCI <- phat + c(-1,1)*t_star*SE_est
    
    p_does[i] <- pCI[1] < p & pCI[2] > p
    phat_does[i] <- phatCI[1] < p & phatCI[2] > p
    that_does[i] <- thatCI[1] < p & thatCI[2] > p
}
```

### Simulation Results

```{r}
mean(p_does)
mean(phat_does)
mean(that_does)
```

Using the population proportion is... worse?\pause

DIY: Change $p$ so that the normal approximation doesn't apply.

## Examples and Cautions

### Example 1 {.t}

It was found that 591 out of 700 people sampled supported a certain political position. Find a 91%CI.

:::notes

Since we have R, let's use it!

Both `prop.test()` and `binom.test()` will give us a CI, with `prop.test()` calculating an approximation using the normal distribution and `binom.test()` calculating the exact value, without approximation. In general, you should always use `binom.test()` for one sample proportions. 

```{r}
binom.test(x = 591, n = 700, conf.level = 0.91)
```
:::



### Example 2 {.t}

It was found that 68 out of 70 people sampled supported a certain political position. Find a 91%CI.\pause

```{r}
n <- 70
phat <- 68/70
se_est <- sqrt(phat*(1-phat)/n)
z_star <- abs(qnorm(0.09/2))

phat + c(-1, 1)*z_star*se_est
```

... so it would be reasonable to say that the popluation proportion is larger than 1???

:::notes
Absolutely not! The normal approximation does not apply here since $n(1 - \hat p) = 70*(1 - 68/70) = 2$, and 2 is less than 10^[Citation needed.]. The normal distribution can only be used when the sample size is large enough!!!

Instead, we can use the exact test. This is much slower to calculate, but for $n = 70$ there's no issue.

```{r}
binom.test(x = 68, n = 70, conf.level = 0.91)
```

Notice that 
:::


### Example 2 {.t}

It was found that 68 out of 70 people sampled supported a certain political position. Find a 91%CI.\pause

```{r}
binom.test(x = 68, n = 70)
```


## Crowdsourced Questions

The following questions are added from the Winter 2024 section of ST231 at Wilfrid Laurier University. The students submitted questions for bonus marks, and I have included them here (with permission, and possibly minor modifications).

1. Assuming all other factors are held constant, how does increasing the number of trials from 100 to 400 affect the width of the confidence interval for a population proportion?
    a. The width of the confidence interval remains unchanged.
    b. The width of the confidence interval decreases, leading to a more precise estimate.
    c. The width of the confidence interval increases, indicating less precision in the estimate.
    d. The change in the width of the confidence interval cannot be determined without knowing the population size.

<details>
<summary>**Solution**</summary>
(b) Increasing the sample size from 100 to 400, while keeping all other factors constant, decreases the width of the confidence interval for a population proportion. This is because the standard error of the proportion decreases as the sample size increases, leading to a more precise estimate of the population proportion.

*****
</details>

















