---
title: "Confidence Intervals for a Proportion"
author: "Devan Becker"
date: "15/07/2020"
output: 
    beamer_presentation:
        colortheme: Western2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

## This is the same as the last video.

- Based on our data, we make an interval that we think describes the population.\newline\pause

- In this case, we just have a different population distribution?

## Assumptions

In stats, assumptions give us power \pause but only if they hold.

\pause

Assumptions for a CI for $p$ are the same as the assumptions for the binomial distribution, with the addition of an SRS.

# The CI for $p$

## Sampling Distribution of $\hat p$

As we saw before the midterm, if the population is $B(n,p)$, then under certain conditions,

$$\hat p \sim N\left(p, \sqrt{\frac{p(1-p)}{n}}\right)$$


## Deja-Vu

Since $\hat p \sim N(p, \sqrt{\frac{p(1-p)}{n}})$, 

$$
\frac{\hat p - p}{\sqrt{p(1-p)/n}} \sim N(0,1)
$$

\pause Again, we can use the form $z = (x-\mu)/\sigma$, but replace $x$, $\mu$, and $\sigma$ with the correct values.\pause

A $(1-\alpha)$CI for $p$ is:

$$
\hat p \pm z^*\sqrt{\frac{p(1-p)}{n}}
$$

## We don't know the variance, why not $t_{n-1}^*$?

- We used $t_{n-1}^*$ because we had to estimate $\sigma$\newline

- There's no $\sigma$ to estimate!\newline

- The variance of the Binomial distribution is entirely determined by $p$!
    - Binom be crazy.


## ... but Devan, we still don't know $p$!


The $(1-\alpha)$CI for $p$ is:

$$
\hat p \pm z^*\sqrt{\frac{p(1-p)}{n}}
$$
which needs $p$ in the second part of the equation.\pause

\quad

Why not just plug in $\hat p$?\pause

Okay fine. \pause

$\sqrt{\hat p(1-\hat p)/n}$ is called the **estimated standard error**, since its the sd of the sampling distribution, but it's based on an estimate.

## Final_Version_V2_Update_LastTry_Srsly.docx.pdf

The $(1-\alpha)$CI for $p$ is:

$$
\hat p \pm z^*\sqrt{\frac{\hat p(1-\hat p)}{n}}
$$

where $z^*$ is chosen such that $P(Z < -z^*) = \alpha/2$.


## Devan Style: Simulation

```{r}
n <- 100
p <- 0.7
SE_true <- sqrt(p*(1-p)/n)
p_does <- c()
phat_does <- c()
that_does <- c()
z_star <- abs(qnorm(0.05/2))
t_star <- abs(qt(0.05/2, df = n-1))
```

## Devan Style: Simulation

```{r}
for(i in 1:10000){
    new_sample <- rbinom(n=1, size=n, prob=p)
    phat <- new_sample/n
    SE_est <- sqrt(phat*(1-phat)/n)
    
    pCI <- phat + c(-1,1)*z_star*SE_true
    phatCI <- phat + c(-1,1)*z_star*SE_est
    thatCI <- phat + c(-1,1)*t_star*SE_est
    
    p_does[i] <- pCI[1] < p & pCI[2] > p
    phat_does[i] <- phatCI[1] < p & phatCI[2] > p
    that_does[i] <- thatCI[1] < p & thatCI[2] > p
}
```

## Simulation Results

```{r}
mean(p_does)
mean(phat_does)
mean(that_does)
```

Using the population proportion is... worse?\pause

DIY: Change $p$ so that the normal approximation doesn't apply.

# Examlpes and Cautions

## Example 1 {.t}

It was found that 591 out of 700 people sampled supported a certain political position. Find a 91%CI.


## Example 2 {.t}

It was found that 68 out of 70 people sampled supported a certain political position. Find a 91%CI.\pause

```{r}
n <- 70
phat <- 68/70
se_est <- sqrt(phat*(1-phat)/n)
z_star <- abs(qnorm(0.09/2))

phat + c(-1, 1)*z_star*se_est
```

... so it would be reasonable to say that the popluation proportion is larger than 1???


## Example 2 {.t}

It was found that 68 out of 70 people sampled supported a certain political position. Find a 91%CI.\pause

```{r}
prop.test(x = 68, n = 70)
```




















